Component response rate variation underlies the stability of complex systems

### 2019-03-02 notebook/ms.tex
Prep the manuscript for both Scientific Reports and arxiv (SI still needs to be fixed in the arxiv tex)

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies stability in large complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Component response rate variation underlies stability in large complex
systems}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important for stability in
large but finite systems, where the probability of stability would
otherwise be negligible. At such extremes of simulated system size, the
largest stable complex systems would be unstable if not for
\(\boldsymbol{Var(\gamma)}\). My results therefore reveal a previously
unconsidered aspect of system stability that is likely to be pervasive
across all realistic complex systems.

@@

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(\mathbf{M}\) matrices at \(\sigma = 0.4\) and
\(C = 1\) across \(S = \{2, 3, ..., 49, 50\}\). One million
\(\mathbf{M}\) were simulated for each \(S\), and the stability of
\(\mathbf{M}\) was assessed given \(\gamma = 1\) versus
\(\gamma \sim \mathcal{U}(0, 2)\). For all \(S > 10\), I found that the
number of stable random systems was higher given \(Var(\gamma)\) than
when \(\gamma = 1\) (Fig. 3; see Supplementary Information for full
table of results), and that the difference between the probabilities of
observing a stable system increased with an increase in \(S\). In other
words, the potential for \(Var(\gamma)\) to drive stability increased
with system complexity and was most relevant for systems on the cusp of
being too complex to be realistically stable. For the highest values of
\(S\), nearly all systems that were stable given \(Var(\gamma)\) would
not have been stable given \(\gamma = 1\).

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a large complex system) is too large to
search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{19}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements (i.e., by
manipulating component response rates rather than interactions between
components).

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that the system is stable and values
of all components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). I want to emphasise that \(\gamma\) is not interpreted as
population density in this analysis, but instead as a fundamental
property of species life history such as expected generation time.
Feasibility was unaffected by \(Var(\gamma)\) and instead occurred with
a fixed probability of \(1/2^{S}\), consistent with a recent proof by
Serván et al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{22}}
(see Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

Here I have shown that the stability of large systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate is more likely to be critical for stability in systems
that are especially large (but finite, see Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}), and it can
ultimately increase the probability that system stability is observed
above that predicted by
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}} classically
derived \(\sigma \sqrt{SC}\) criterion. The logic outlined here is
general, and potentially applies to any complex system in which
individual system components can vary in their reaction rates to system
perturbation.

It is important to emphasise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rate to a particular system does not necessarily increase the
probability that the system will be stable (at high \(S\), the opposite
is true; see Supplementary Information). Rather, highly complex systems
that are observed to be stable are more likely to have varying component
response rates, and for this variation to be critical to their stability
(Fig. 3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the community matrix \(\mathbf{M}\) to become negative, but might
also increase the real components of eigenvalues. In this way, my focus
is distinct from that of Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results are not inconsistent with this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity, \(\sigma\sqrt{SC}\) (see Supplemental Information for
results simulated across \(\sigma\) and \(C\). Overall, my results show
that component response rate variation increases the upper bound of
complexity at which stability can be realistically observed, meaning
that highly complex systems are more likely than not to vary in their
component response rates, and for this variation to be critical for
system stability.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of
species\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Servan2018}{22}}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the feasibility of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical
systems\textsuperscript{\protect\hyperlink{ref-Patel2018}{15}}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a vector \(\gamma\) has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}},
but \(\gamma\) has not been interpreted as rates of response of
individual system components to perturbation. My model focuses on
component response rates for systems of a finite size, in which
complexity is high but not yet high enough to make the probability of
stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 3). Overall, I suggest that
variation in component response rate might therefore be critical for
maintaining stability in many highly complex empirical systems. These
results are broadly applicable to understanding stability of complex
networks across the physical, life, and social sciences.

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) modulates the values of
components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{23}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) row means that is
attributable to \(\mathbf{A}\) is small relative to that attributable to
\(\mathbf{\gamma}\), especially at high \(S\). Variation in
\(\mathbf{\gamma}\) specifically isolates the effects of differing
component response rates, hence causing differences in expected
\(\mathbf{M}\) row means.

@@

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M}\) (\(\sigma = 0.4\), \(C = 1\)), where
\(\mathbf{M = \gamma A}\). The genetic algorithm was initialised with a
population of 1000 different \(\mathbf{\gamma}\) vectors with elements
sampled i.i.d from \(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis
was performed on the \(\mathbf{M}\) resulting from each
\(\mathbf{\gamma}\) vector, and the 20 \(\mathbf{\gamma}\) vectors
resulting in \(\mathbf{M}\) with the lowest
\(\max\left(\Re(\lambda)\right)\) each produced 50 clonal offspring with
subsequent random mutation and crossover between the resulting new
generation of 1000 \(\mathbf{\gamma}\) vectors. Mutation of each
\(\gamma_{i}\) in a \(\mathbf{\gamma}\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100 \(\mathbf{\gamma}\)
vectors paired in each generation; vectors were randomly sampled with
replacement among but not within sets. Vector pairs selected for
crossover swapped all elements between and including two \(\gamma_{i}\)
randomly selected with replacement (this allowed for reversal of vector
element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\mathbf{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\max\left(\Re(\lambda)\right)\) for
\(\mathbf{M}\)).

\textbf{System feasibility}. Dougoud et
al.\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}} identify
the following feasibility criteria for ecological systems characterised
by \(S\) interacting species with varying densities in a classical
Lotka-Volterra model,

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.726 and 0.742 for a and b,
respectively.

@@

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies the stability of complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Component response rate variation underlies the stability of complex
systems}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I show that when component response rates vary, the
potential for system stability is markedly increased. Variation in
\(\boldsymbol{\gamma}\) becomes increasingly important for stability in
large but finite systems, where the probability of stability would
otherwise be negligible. At such extremes of simulated system
complexity, the largest stable complex systems would be unstable if not
for \(\boldsymbol{Var(\gamma)}\). My results therefore reveal a
previously unconsidered aspect of system stability that is likely to be
pervasive across all realistic complex systems.

@@

\textbf{Simulation of random \(\mathbf{M}\) across \(\mathbf{S}\)}. To
investigate the effect of \(Var(\gamma)\) on stability across systems of
varying complexity, I simulated random \(\mathbf{M}\) matrices at
\(\sigma = 0.4\) and \(C = 1\) across \(S = \{2, 3, ..., 49, 50\}\). One
million \(\mathbf{M}\) were simulated for each \(S\), and the stability
of \(\mathbf{M}\) was assessed given \(\gamma = 1\) versus
\(\gamma \sim \mathcal{U}(0, 2)\). For all \(S > 10\), I found that the
number of stable random systems was higher given \(Var(\gamma)\) than
when \(\gamma = 1\) (Fig. 3; see Supplementary Information for full
table of results), and that the difference between the probabilities of
observing a stable system increased with an increase in \(S\). In other
words, the potential for \(Var(\gamma)\) to affect stability increased
with system complexity and was most relevant for systems on the cusp of
being too complex to be realistically stable. For the highest values of
\(S\), nearly all systems that were stable given \(Var(\gamma)\) would
not have been stable given \(\gamma = 1\).

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{19}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements (i.e., by
manipulating component response rates rather than interactions between
components).

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that values of all components are
positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Song2018}{21}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). I emphasise that \(\gamma\) is not interpreted as
population density in this analysis, but instead as a fundamental
property of species life history such as expected generation time.
Feasibility was unaffected by \(Var(\gamma)\) and instead occurred with
a fixed probability of \(1/2^{S}\), consistent with a recent proof by
Serván et al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{22}}
(see Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

I have shown that the stability of complex systems might often be
contigent upon variation in the response rates of their individual
components, meaning that factors such as rate of trait evolution (in
biological networks), transaction speed (in economic networks), or
communication speed (in social networks) need to be considered when
investigating the stability of complex systems. Variation in component
response rate is more likely to be critical for stability in systems
that are especially complex, and it can ultimately increase the
probability that system stability is observed above that predicted by
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}} classically
derived \(\sigma \sqrt{SC}\) criterion. The logic outlined here is
general, and potentially applies to any complex system in which
individual system components can vary in their reaction rates to system
perturbation.

It is important to recognise that variation in component response rate
is not stabilising per se; that is, adding variation in component
response rates to a particular system does not increase the probability
that the system will be stable. Rather, highly complex systems that are
observed to be stable are more likely to have varying component response
rates, and for this variation to be critical to their stability (Fig.
3). This is caused by the shift to a non-uniform distribution of
eigenvalues that occurs by introducing \(Var(\gamma)\) (Fig. 1b, 2b),
which can sometimes cause all of the real components of the eigenvalues
of the system matrix to become negative, but might also increase the
real components of eigenvalues.

My focus is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity, \(\sigma\sqrt{SC}\) (see Supplemental Information for
results simulated across \(\sigma\) and \(C\)). Overall, I show that
component response rate variation increases the upper bound of
complexity at which stability can be realistically observed, meaning
that highly complex systems are more likely than not to vary in their
component response rates, and for this variation to be critical for
system stability.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of
species\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{18},\protect\hyperlink{ref-Dougoud2018}{20},\protect\hyperlink{ref-Servan2018}{22}}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the stability of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical
systems\textsuperscript{\protect\hyperlink{ref-Patel2018}{15}}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a vector \(\gamma\) has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{16}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17},\protect\hyperlink{ref-Stone2017}{18}},
but \(\gamma\) has not been interpreted as rates of response of
individual system components to perturbation. My model focuses on
component response rates for systems of a finite size, in which
complexity is high but not yet high enough to make the probability of
stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 3). Overall, I suggest that
variation in component response rate might therefore be critical for
maintaining stability in many highly complex empirical systems. These
results are broadly applicable for understanding the stability of
complex networks across the physical, life, and social sciences.

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) defines the values of system
components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{16},\protect\hyperlink{ref-May1973}{23}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in \(\mathbf{M}\) row
means that is attributable to \(\mathbf{A}\) is small relative to that
attributable to \(\mathbf{\gamma}\), especially at high \(S\). Variation
in \(\mathbf{\gamma}\) specifically isolates the effects of differing
component response rates, hence causing differences in expected
\(\mathbf{M}\) row means.

@@

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M}\) (\(\sigma = 0.4\), \(C = 1\)). The
genetic algorithm was initialised with a population of 1000 different
\(\mathbf{\gamma}\) vectors with elements sampled i.i.d from
\(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis was performed on
the \(\mathbf{M}\) resulting from each \(\mathbf{\gamma}\) vector, and
the 20 \(\mathbf{\gamma}\) vectors resulting in \(\mathbf{M}\) with the
lowest \(\max\left(\Re(\lambda)\right)\) each produced 50 clonal
offspring with subsequent random mutation and crossover between the
resulting new generation of 1000 \(\mathbf{\gamma}\) vectors. Mutation
of each \(\gamma_{i}\) in a \(\mathbf{\gamma}\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100 \(\mathbf{\gamma}\)
vectors paired in each generation; vectors were randomly sampled with
replacement among but not within sets. Vector pairs selected for
crossover swapped all elements between and including two \(\gamma_{i}\)
randomly selected with replacement (this allowed for reversal of vector
element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\mathbf{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\max\left(\Re(\lambda)\right)\) for
\(\mathbf{M}\)).

\textbf{System feasibility}. Dougoud et
al.\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}} identify
the following feasibility criteria for ecological systems characterised
by \(S\) interacting species with varying densities in a generalised
Lotka-Volterra model,

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.724 and 0.742 for a and b,
respectively.

@@

\includegraphics{ms_files/figure-latex/unnamed-chunk-16-1.pdf}
